Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-11-07 response_to_reviewers.Rmd

---
title: "Response to Reviewers"
date: "2022-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major comments from Reviewer 1

> Overall, I thought the authors were successful in introducing the big ideas in sports
analytics at a general and non-technical level, and many interesting aspects of the literature
were discussed. However, I think the second part of the title ”Statistical Tools for Their
Investigation” was not properly explored. Several packages are mentioned throughout, and
a brief sketch of a possible analysis workflow is given in Section 5, but I think it lacks sufficient
detail. Beyond that, I think there are some extraneous figures and possible improvements
with the overall flow.

Our revision includes a significant bolstering of Sections 4, 5, and 6. In particular, we have added a subsection on Elo ratings, added a subsection on baseball tools, expanded and linked the discussion of graphical tools, and replaced some weak figures with more relevant substitutes.

@@

## Major comments from Reviewer 2

> While Sections 2 and 3 are well done and complete, Sections 4 and 5 are weak and seem incomplete and unpolished.

As noted above, Sections 4 and 5 are significantly stronger in this revision.

@@

> I think this section could use a table of some sort, perhaps something that summarizes the CRAN task view at a high level. Another potential would be a figure which shows a typical workflow of sports analytics as explained in the text.

We have added a table summarizing the sport-specific packages that are available on CRAN.

> Figure 6 doesn't add much to the text; I don't think the reader better understands sports analytics by visualizing many team colors. Instead, it would be better to show  how the theming can lead to a more impactful graphic for sports, i.e. show how that package might be used in a real-world setting.

We thank the reviewer for this excellent suggestion. We have replaced Figure 6 with a new figure that illustrates in the `retrosheet` and `teamcolors` package can be combined to produce a meaningful data graphic. In particular, we note how the use of team colors clarifies what would otherwise be a messy jumble of indistinguishable lines.

@@

> 12, 23: This is not much of a case study.  Perhaps some additional information on the data available, ways it could be used to calculate game state expected valued, win probability, player ratings….

We have connected the case study in chess to the new subsection on Elo ratings and the subsequent subsection on graphical tools. We hope that these explicit connections to additional content will make this subsection feel more substantial. 
In addition, we have added new subsection to this Section with a different case study in baseball about the evolution of data, tools, and statistical methodologies. We hope that this additional content will help to flesh out the value of these tools.

> 12, 37: Theming *teamcolors* is a great *R* package, not sure it is part of the "big ideas" in sports analytics.  A single line in the first paragraph of this section would seem to be better.

We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection. We also replaced the previous Figure 6 with a new Figure 6 that illustrates how the use of team colors is particularly impactful for sports-releated data graphics.

### 2022-11-07 sports_analytics.Rmd
updated response

Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the longest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact. 
In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact.

@@

```{r mlb_standings, fig.cap="The progression of National Leauge team standings during the 2021 Major League Baseball season. Note how the use of team colors makes it possible to untangle what would otherwise be a messy jumble of indistinguishable lines. Data provided by \\pkg{retrosheet} and colors provided by \\pkg{teamcolors}."}
ggplot(
    games_2021, 
    aes(x = cum_G, y = G_over_500, color = team_id)
  ) + 
  geom_hline(yintercept = 18, color = "darkgray", lty = 2) + 
  geom_line() +  
  ggrepel::geom_label_repel(
    data = filter(games_2021, last_game), 
    aes(label = team_id)
  ) + 
  scale_color_manual(
    guide = "none", name = "Team",
    values = mlb_pal
  ) +
  scale_y_continuous("Games above .500") + 
  scale_x_continuous("Cumulative Games Played") + 
  labs(title = "Daily MLB Standings, 2021", 
       caption = "Source: Retrosheet")
```

>>>>

Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the longest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact. 
In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].

@@

```{r mlb_standings, fig.cap="The progression of National Leauge team standings during the 2021 Major League Baseball season. Note how the use of team colors makes it possible to untangle what would otherwise be a messy jumble of indistinguishable lines. Data provided by \\pkg{retrosheet} and colors provided by \\pkg{teamcolors}."}
ggplot(
    games_2021, 
    aes(x = cum_G, y = G_over_500, color = team_id)
  ) + 
  geom_hline(yintercept = 18, color = "darkgray", lty = 2) + 
  geom_line() +  
  ggrepel::geom_label_repel(
    data = filter(games_2021, last_game), 
    aes(label = team_id)
  ) + 
  annotate("text", x = 2, y = 22, label = "90 wins", color = "darkgray", hjust = "left") +
  scale_color_manual(
    guide = "none", name = "Team",
    values = mlb_pal
  ) +
  scale_y_continuous("Games above .500") + 
  scale_x_continuous("Cumulative Games Played") + 
  labs(title = "Daily MLB Standings, 2021", 
       caption = "Source: Retrosheet")
```
